---
title: "Module 19 - Association mapping: GWAS & sequencing data"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r echo = FALSE, message=FALSE, warning=FALSE}
require(dplyr)
print.link <- function(name, link) {
  if(link != "") { 
    paste0("[", name, "](", link, ")")
  } else paste0("[", name, "]")
}
```

Instructors: Joelle Mbatchou & Loic Yengo

You will find on this website link to lecture slides, exercises including some solutions, and link to the class video recordings.

The official SISG web page can be found [here](https://sisg.biosciences.gatech.edu/sisg-module-19/) (requires login) – course evaluations and certificates of completion are available via this page.


## Session Format
The module has 10 sessions, each of 90 minutes. The standard format for a session is approximately:

* 45 minutes of lecture material that will be recorded via Zoom live and posted at the end of the day
* 30 minutes of exercises for you to try, with small-group "breakout" Zoom sessions available, attended by other class participants, and Teaching Assistants
* 15 minute discussion of exercises, where the instructors will present possible solutions and answer questions

Each session builds on previous sessions so you will get maximum benefit by attending every session live and in sequence.

The course schedule is below.


## Pre-requisites

We will be running exercises during the course using various tools, follow the instructions below to get setup on your machine.

**Prior to the module**, please install working and latest versions of:

* R (Version 4.1+): 
  <a href="https://www.r-project.org/" target="_blank">https://www.r-project.org/</a>
* RStudio (Version 1.4+) : 
  <a href="https://rstudio.com/products/rstudio/download/" target="_blank">https://rstudio.com/products/rstudio/download/</a>
* PLINK 1.9 : 
  <a href="https://www.cog-genomics.org/plink2/" target="_blank">https://www.cog-genomics.org/plink2/</a>
* PLINK 2.0 : 
  <a href="https://www.cog-genomics.org/plink/2.0/" target="_blank">https://www.cog-genomics.org/plink/2.0/</a>
* REGENIE : 
  <a href="https://github.com/rgcgithub/regenie/releases/" target="_blank">https://github.com/rgcgithub/regenie/releases/</a>
* GCTA : 
  <a href="https://yanglab.westlake.edu.cn/software/gcta/#Download" target="_blank">https://yanglab.westlake.edu.cn/software/gcta/#Download</a>
* GCTB : 
  <a href="https://cnsgenomics.com/software/gctb/#Download" target="_blank">https://cnsgenomics.com/software/gctb/#Download</a>

All are free software. For those on Windows, we recommend WSL ([Windows Subsystem for Linux](https://learn.microsoft.com/en-us/windows/wsl/install)) to run the tools above.

Please be sure to install some necessary R packages with:
```{r install-pkgs, eval = FALSE}
install.packages("qqman")
install.packages("bigsnpr")
install.packages("hexbin")
install.packages("devtools")
install.packages("tidyverse")
install.packages("cowplot")
install.packages("BEDMatrix")
install.packages("SKAT")
devtools::install_github("yaowuliu/ACAT")
```

## Schedule

All times listed below for the schedule are Eastern Daylight Time (EDT, Atlanta time).

```{r wed-sched, echo = FALSE, results = "asis"}
data.frame(
  Time = c("1:30pm-3pm", "3:30pm-5pm"),
  Session = 1:2,
  Topic = c(
    "Introduction, Case Control Association Testing & Association Testing with Quantitative Traits",
    "Introduction to the PLINK Software for GWAS & Population Structure Inference"
    ),
  Lecture = c(
    paste0(print.link("Intro", ""), ", ", print.link("Lecture", "")),
    paste0(print.link("Lecture", ""))
      ),
  "Exercises" = c(
    paste0(print.link("Exercises", "Session01_practical.html")),
    paste0(print.link("Exercises", "Session02_practical.html"))
  ),
  "Key" = c(
paste0(print.link("HTML", ""), ", ", print.link("Rmd", "")),
paste0(print.link("HTML", ""), ", ", print.link("Rmd", ""))
  )
 , check.names = FALSE ) %>%
knitr::kable("html", escape = FALSE, 
  caption = "Wednesday, June 12th", 
  align = c('l', 'c', 'l', 'l', 'l')
) %>%
kableExtra::kable_styling(bootstrap_options = c("hover", "condensed","striped"))
```

```{r thu-sched, echo = FALSE, results = "asis"}
data.frame(
  Time = c("8:30am-10am", "10:30am-12pm", "1:30pm-3pm", "3:30pm-5pm"),
  Session = 2 + 1:4,
  Topic = c(
    "GWAS in Samples with Structure & Introduction to the REGENIE Software",
    "Power, Sample Size & Design Considerations",
    "Downstream analyses: Conditional Analyses, Colocalization, Fine-mapping, Gene-based & Pathways Analysis",
    "Downstream analyses: Polygenic Risk Scores"
    ),
  Lecture = c(
    paste0(print.link("Lecture", "")),
    paste0(print.link("Lecture", "")),
    paste0(print.link("Lecture", "")),
    paste0(print.link("Lecture", ""))
      ),
  "Exercises" = c(
    paste0(print.link("Exercises", "Session03_practical.html")),
    paste0(print.link("Exercises", "SISGM15_prac4.html")),
    paste0(print.link("Exercises", "SISGM15_prac5.html")),
    paste0(print.link("Exercises", "SISGM15_prac6.html"))
  ),
  "Key" = c(
paste0(print.link("HTML", ""), ", ", print.link("Rmd", "")),
paste0(print.link("HTML", ""), ", ", print.link("Rmd", "")),
paste0(print.link("HTML", ""), ", ", print.link("Rmd", "")),
paste0(print.link("HTML", ""), ", ", print.link("Rmd", ""))
  )
  , check.names = FALSE) %>%
knitr::kable("html", escape = FALSE, 
  caption = "Thursday, June 13th", 
  align = c('l', 'c', 'l', 'l', 'l')
) %>%
kableExtra::kable_styling(bootstrap_options = c("hover", "condensed","striped"))
```

```{r fri-sched, echo = FALSE, results = "asis"}
data.frame(
  Time = c("8:30am-10am", "10:30am-12pm", "1:30pm-3pm", "3:30pm-5pm"),
  Session = 6 + 1:4,
  Topic = c(
    "Rare Variant Analysis: Collapsing Tests, Kernel (Variance Component) Tests and Omnibus Tests",
    "Emerging issues showcasing ongoing research",
    "Beyond traditional GWAS: Variance QTL, Interaction Testing, GWAX, Time-to-event, Multi-trait analysis",
    "Emerging issues showcasing ongoing research"
    ),
  Lecture = c(
    paste0(print.link("Lecture", "")),
    "Lecture" 
      ),
  "Exercises" = c(
    paste0(print.link("Exercises", "Session07_practical.html")),
    paste0(print.link("Exercises", "Session08_practical.html")),
    paste0(print.link("Exercises", "SISGM15_prac9.html")),
    " "
  ),
  "Key" = c(
paste0(print.link("HTML", ""), ", ", print.link("Rmd", "")),
paste0(print.link("HTML", ""), ", ", print.link("Rmd", "")),
   " ",# paste0(print.link("HTML", ""), ", ", print.link("Rmd", "")),
    " "
  )
  , check.names = FALSE) %>%
knitr::kable("html", escape = FALSE, 
  caption = "Friday, June 14th", 
  align = c('l', 'c', 'l', 'l', 'l')
) %>%
kableExtra::kable_styling(bootstrap_options = c("hover", "condensed","striped"))
```


## Datasets

All the data sets used can be downloaded from the [Box folder](https://gatech.box.com/s/l9rfqhs8v4o9j8bm9qg6conm4zax7rnp) for the class.

## Additional resources

* [2016 TPB Paper](https://faculty.washington.edu/tathornt/sisg/Zheng_Weir_Eigenanalysis_TPB_2016.pdf) on PCA for Population Structure Inference and Admixture Estimation
* [HapMap MXL Individual Ancestry Figure](https://faculty.washington.edu/tathornt/sisg/HapMapMXL_Ancestry_Supervised.pdf) (Supervised Learning)
* If you are new to R, you may find the following material helpful:
  * [Introduction to R materials](https://faculty.washington.edu/kenrice/rintro/) from a past SISG Module
  * [Graphics with ggplot2](https://ggplot2.tidyverse.org/)
  * [Data manipulation with dplyr](http://dplyr.tidyverse.org/)

